---
title: "Conditional Copula"
output: pdf_document
---

```{r include=FALSE}
library("ggplot2")
library("mixtools")
library("mvtnorm")
```

## Example of a Conditional Copula


\def\mb #1{\boldsymbol{#1}}

Consider the two-dimensional random vector $(X, Y)$ having joint conditional distribution
$$
f(x,y|z)=f_X(x)f_Y(y)c(F_X(x),F_Y(y)\mid z)
$$
where each marginal is a mixture of univariate normal random variables.
$$
f_X(x) = f_Y(x) = \sum_{r=1}^{R}\pi_r\phi(x\mid \mu_r, \sigma_r)
$$
We chose the following settings and parameter values: $f_X = f_Y$, $R = 6$, $\mu_{1:6} = (-9, -5.4, -1.8, 1.8, 5.4, 9)$, and $\sigma_r = 1/\sqrt{10}$, $\pi_r = 1/R$.

and the copula is
$$
C(u,v \mid x )=u^{1-\alpha(x)}v^{1-\beta(x)}[u^{-\theta\alpha(x)}+v^{-\theta\beta(x)}-1]^{-1/\theta}
$$

where 

- $\theta=20$

- $$
  \alpha(x)=\frac{2}{3}-\frac{1}{4}\times\frac{1}{\exp(x)+1}
  $$

  

- $$
  \beta(x)=\frac{1}{2}+\frac{1}{4}\times\frac{\exp(x)}{\exp(x)+1}
  $$



Draw a sample when $x\in (-4,4)$

```{r}

set.seed(100)

R=6
mu=c(-9, -5.4, -1.8, 1.8, 5.4, 9)
sigmar=rep(1/sqrt(10), R)
pi=rep(1/R, R)
theta=40

# evaluate the function at the point x, where the components 
# of the mixture have weights w, means stored in u, and std deviations
# stored in s - all must have the same length.
F = function(x,w,u,s) sum( w*pnorm(x,mean=u,sd=s) )

# provide an initial bracket for the quantile. default is c(-1000,1000). 
F_inv = function(p,w,u,s,br=c(-1000,1000))
{
  G = function(x) F(x,w,u,s) - p
  return( uniroot(G,br)$root ) 
}


l=5
#function alpha(x)
a<-function(x){
  2/3-1/4*(exp(l*x)/(exp(l*x)+1))
}
#function beta(x)
b<-function(x){
  1/2+1/4*(exp(l*x)/(exp(l*x)+1))
}

n=1500
V=runif(n)
W=runif(n)

X=seq(-4,4, length.out = n)

C <- function(u) {
  (1-beta)*u^(1-alpha)*v^(-beta)*(u^(-theta*alpha)+v^(-theta*beta)-1)^(-1/theta)+
    beta*u^(1-alpha)*v^(-beta*(1+theta))*(u^(-theta*alpha)+v^(-theta*beta)-1)^(-1/theta-1)-w
}

U=rep(0,n)
for (i in 1:n) {
  v=V[i]
  w=W[i]
  alpha=a(X[i])
  beta=b(X[i])
  U[i]=uniroot(C, lower = 0, upper = 1)$root
}
```


```{r}
par(mfrow=c(1,3))
plot(U[1:(1/3*n)], V[1:(1/3*n)], xlab = 'U', ylab='V', main = '-4<x<-1.3')
plot(U[(1/3*n):(2/3*n)], V[(1/3*n):(2/3*n)],xlab = 'U', ylab='V', main = '-1.3<x<1.3')
plot(U[(2/3*n):n], V[(2/3*n):n], xlab = 'U', ylab='V', main = '1.3<x<4')
```


```{r}
X1=sapply(V, F_inv, w=pi, u=mu, s=sigmar)
X2=sapply(U, F_inv, w=pi, u=mu, s=sigmar)

par(mfrow=c(1,3))
plot(X1[1:(1/3*n)], X2[1:(1/3*n)],  xlab = 'X', ylab='Y', main ='-4<x<-1.3')
plot(X1[(1/3*n):(2/3*n)], X2[(1/3*n):(2/3*n)], xlab = 'X', ylab='Y', main =  '-1.3<x<1.3')
plot(X1[(2/3*n):n], X2[(2/3*n):n], xlab = 'X', ylab='Y', main = '1.3<x<4')
```


# Another Example of Conditional Copula

Notation:

- $\Phi$ is the cdf of the standard normal distribution. 
- $\phi$ is the pdf of standard normal distribution.
- $\phi_2$ is the pdf of bivariate normal.

Then they have
$$
p(u_1,u_2 \mid \mu, \sigma)=\frac{\phi_2(\Phi^{-1}(u_1), \Phi^{-1}(u_2)  \mid\mb\mu, \sigma I_2)}{\phi(\Phi^{-1}(u_1))\phi(\Phi^{-1}(u_2))}
$$
**A mixture of $p(\mb u)$ can be used to estimate any arbitrarily continuous density on $(0,1)^2$.**
$$
\tilde{c}(u_1,u_2 \mid \mb \pi, \mb \mu, \mb \sigma) =\sum^J_{j=1} \pi_j \frac{\phi_2(\Phi^{-1}(u_1),\Phi^{-1}(u_2)  \mid \mb\mu_j, \sigma_j I_2)}{\phi(\Phi^{-1}(u_1))\phi(\Phi^{-1}(u_2))}
$$
where $\pi_j$s are mixing proportions.



**When $\pi_j$ and $\mu_j$  are conditioned on covariates:**

- $$
  \pi_j(x_i)=\frac{\exp(\beta_j^T x_i)}{\sum^J_{j=1} \exp(\beta_j^Tx_i)}\\
  $$

- $$
  \mb\mu_{j}=
  \begin{bmatrix}
  \lambda_{j1}^Tx_i \\
  \lambda_{j2}^Tx_i 
  \end{bmatrix}
  $$



We give an example  of a mixture with two components

- $$
  \beta=
  \begin{bmatrix}
  0.1 \\
  0.3
  \end{bmatrix}
  $$

  

- $$
  \lambda_{11}=
  \begin{bmatrix}
  0.1 \\
  0.3
  \end{bmatrix}
 $$
 
 $$
  \lambda_{12}=
  \begin{bmatrix}
  0.1 \\
  0.4
  \end{bmatrix}
  $$

  $$
  \lambda_{21}=
  \begin{bmatrix}
  0.3\\
  0.1 
  \end{bmatrix}
  $$
  
  $$
  \lambda_{22}=
  \begin{bmatrix}
  0.4\\
  0.1 
  \end{bmatrix}
  $$

- $\sigma_1=0.5$ and $\sigma_2=1$.

Draw a random sample from above model when $X\in(0, 5)$.

```{r}
n=1500
X=seq(0,5, length.out = n)
X=cbind(rep(1, n), X)
beta1=c(0.1, 0.3)

pi=cbind(exp(X%*%beta1)/(exp(X%*%beta1)+1), 1/(exp(X%*%beta1)+1))

h=t(apply(pi, 1,  rmultinom, n=1, size=1))

Z1=matrix(rep(0, 2*n), nrow = n)

lambda11=c(0.1,0.3)
lambda12=c(0.1,0.4)
sigma=0.5
I=matrix(c(1,0,0,1), nrow = 2)
mu1=cbind((X%*%lambda11)[,1], (X%*%lambda12)[,1])

for (i in 1:n) {
  Z1[i, ]=rmvnorm(1,mu1[i,], sigma*I)
}


Z2=matrix(rep(0, 2*n), nrow = n)

lambda21=c(0.3, 0.1)
lambda22=c(0.4, 0.1)
sigma2=1
I=matrix(c(1,0,0,1), nrow = 2)
mu2=cbind((X%*%lambda21)[,1], (X%*%lambda22)[,1])

for (i in 1:n) {
  Z2[i, ]=rmvnorm(1,mu2[i,], sigma2*I)
}
Z=matrix(rep(0, 2*n), nrow = n)
for (i in 1:n) {
Z[i,]=Z1[i,]*h[i,1]+Z2[i,]*h[i,2]
}

u=t(apply(Z, 1, pnorm))
U=u[,1]
V=u[,2]

par(mfrow=c(1,3))
plot(U[1:(1/3*n)], V[1:(1/3*n)], xlab = 'U', ylab='V', main = '0<x<-1.5')
plot(U[(1/3*n):(2/3*n)], V[(1/3*n):(2/3*n)],xlab = 'U', ylab='V', main = '1.5<x<3')
plot(U[(2/3*n):n], V[(2/3*n):n], xlab = 'U', ylab='V', main = '3<x<5')
```


```{r}
set.seed(100)
R=6
mu=c(-9, -5.4, -1.8, 1.8, 5.4, 9)
sigmar=rep(1/sqrt(10), R)
pi=rep(1/R, R)
X1=sapply(V, F_inv, w=pi, u=mu, s=sigmar)
X2=sapply(U, F_inv, w=pi, u=mu, s=sigmar)

par(mfrow=c(1,3))
plot(X1[1:(1/3*n)], X2[1:(1/3*n)],  xlab = 'X', ylab='Y', main = '0<x<-1.5')
plot(X1[(1/3*n):(2/3*n)], X2[(1/3*n):(2/3*n)], xlab = 'X', ylab='Y', main = '1.5<x<3')
plot(X1[(2/3*n):n], X2[(2/3*n):n], xlab = 'X', ylab='Y', main = '3<x<5')
```